\documentclass[a4paper,11pt]{article}

\usepackage{xspace, hyperref, listings}

% The next three lines deal with the underscore character
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{textcomp}


% This line needs package svn-multi
% \svnid{$Id$}

\pagestyle{plain}
\linespread{1.6}
\textwidth        16 cm
\textheight       22 cm
\oddsidemargin     0 mm
\evensidemargin    0 mm
\topmargin       -13 mm

% Prevent indenting paragraphs
\setlength\parindent{0pt}

\lstset{basicstyle=\ttfamily, frame=shadowbox, breaklines=true, breakatwhitespace=true}

\newcommand{\modulename}{ContinuumElectrostatics\xspace}
\newcommand{\pka}{$\mathrm{p}K_{\mathrm{a}}$\xspace}


%=================================================
\begin{document}

\begin{center}
{\LARGE \bf \modulename}

{\large
A pDynamo module for protonation state calculations

\vspace{1.0cm}
\underline{Mikolaj Feliks}\\
Institut de Biologie Structurale, Grenoble\\

Last updated: \today %Code revision: \svnrev
}

\vspace{0.5cm}

\end{center}


%=================================================
\section{Introduction}
%=================================================
Proteins contain residues, cofactors and ligands that bind or release protons
depending on the current pH and the interactions with their molecular
environment.
%
These titratable residues, cofactors and ligands will be referred to as sites.
%
Each site exists in at least two charge forms, usually the protonated form and 
the deprotonated form.
%
Different charge forms of a site are called instances.
%
The titration of proteins is difficult to study experimentally because
the available methods, such as calorimetry, cannot determine protonation states
of individual sites.
%
The knowledge of these individual protonation states is crucial for
understanding of many important processes, for example enzyme catalysis\cite{Bombarda2010}.


The \modulename module extends the pDynamo software library by 
Martin Field\cite{pDynamo_Field2008}  with a Poisson-Boltz\-mann continuum electrostatic 
model that allows for calculations of protonation states of individual 
sites\cite{Bashford1992,Ullmann1999,Bombarda2006}.
%
The module connects pDynamo to the external solver of the Poisson-Boltzmann 
equation, MEAD\cite{Bashford1997}.
%
MEAD was developed by Donald Bashford and 
later extended by Timm Essigke\cite{Essigke_PhD} and Thomas Ullmann\cite{Thomas_PhD}.
%
The electrostatic energy terms obtained by MEAD can be used to calculate
energies of all possible protonation states of the protein of interest.
%
However, analytic evaluation of protonation state energies is only possible
for proteins with usually less than 20 titratable sites.
%
The \modulename module provides an interface to the GMCT\cite{Ullmann2012,Thomas_PhD}
program by Matthias Ullmann and Thomas Ullmann that can be used to sample
protonation state energies using a Monte Carlo method.
%
The GMCT interface makes it possible to study the titration of larger proteins.


%=================================================
\section{Copying}
%=================================================
The module is distributed under the CeCILL Free Software License, which is
a French equivalent of the GNU General Public License.
%
For details, see the files \texttt{Licence\_CeCILL\_V2-en.txt} (or
\texttt{Licence\_CeCILL\_V2-fr.txt} for the French version).


%=================================================
\section{Goals of the \modulename module}
%=================================================
I wrote the \modulename module primarily as a pretext to learn how
the Poisson-Boltzmann model works in detail.
%
I used this model very often during my studies on enzyme catalysis but never had
time nor pressure to learn the details of the theory that was behind.
%
I also wanted to better explore the pDynamo library, understand its programming
concepts and finally extend it with something useful.
%
Last but not least, I saw some room for improvement in the previously used
tools and scripts.

The \modulename module is similar in behavior to the \texttt{multiflex2qmpb.pl}
program, which is part of the QMPB package written by Timm Essigke\cite{Essigke_PhD}.
%
The approach taken here is most compatible with the original approach
by Donald Bashford\cite{Bashford1992}.
%
The key novelty is the improved treatment of multiprotic sites, 
such as histidine.
%
Also, the calculations with the \makebox{\modulename} module are considerably faster 
than previously because of the \makebox{parallelization}.

% Talk about non-binary state vector and different energy reference (fully deprotonated system)
%
% For the calculated thermodynamic cycle, see Fig. 3.8, p. 82 in Timm's thesis.


%=================================================
\section{Installation and configuration}
%=================================================
Before the installation of the \modulename module, it is necessary
to have:
\begin{itemize}
  \setlength{\itemsep}{2pt}
  \item pDynamo 1.8.0
  \item Python 2.7
  \item Cython 0.15.1 (optionally, for recompiling Pyrex sources)
  \item GCC (any version should be fine)
  \item Extended MEAD 2.3.0
  \item GMCT 1.2.3
\end{itemize}
%
Extended MEAD and GMCT can be found on the website of Thomas Ullmann:

\url{http://www.bisb.uni-bayreuth.de/People/ullmannt/index.php?name=software}

Download the two packages and follow their respective installation
instructions.
%
The \modulename module requires for its functioning two programs
from the MEAD package, namely \texttt{my\_2diel\_solver} and \texttt{my\_3diel\_solver},
and the GMCT's main program, \texttt{gmct}.

\bigskip
In the next step, check out the latest source code of the module.
%
Note that for checking out the source code you should have Subversion installed
as well.

{\footnotesize \begin{lstlisting}
svn checkout http://pdynamo-extensions.googlecode.com/svn/trunk/ContinuumElectrostatics/
\end{lstlisting} }

\bigskip
Some parts of the module implementing the state vector and the calculation of the microstate energy
are written in C and therefore have to be compiled before use. 

\bigskip
Start from going to the directory \texttt{extensions/csource/}.
%
Change the uppermost line in the \texttt{Makefile}, starting from "PDYNAMO\_PCORE".
%
This line defines the location of the pCore module of your pDynamo installation.
%
After editing, close the file and run "make" to generate an object file.

\bigskip
Go to the directory \texttt{extensions/pyrex/}.
%
Adjust the line starting from \makebox{"PDYNAMO\_PCORE"} in the \texttt{Makefile}.
%
After closing the file, run "make".
%
If you also want to recompile the Pyrex sources, run "make all" instead.
%
In case the Pyrex compilation fails, it may be necessary to manually specify the location of the pCore module in the {\texttt{pyrex\_compile.py} script.
%
Finally, run "make install" to install the compiled library.

\bigskip
At this point, the installation is complete.

\bigskip
Before using the module, the environment variable \texttt{PDYNAMO\_CONTINUUMELECTROSTATICS} 
should be set to the module's root directory.
%
This directory should be also added to the \texttt{PYTHONPATH} variable. 
%
This can be done in the following way (in Bash):

\newpage
{\footnotesize \begin{lstlisting}
export PDYNAMO_CONTINUUMELECTROSTATICS=/home/mikolaj/devel/ContinuumElectrostatics
export PYTHONPATH=$PYTHONPATH:$PDYNAMO_CONTINUUMELECTROSTATICS
\end{lstlisting} }


%=================================================
\section{Usage}
%=================================================
After the installation, it may be worth looking at some of the test cases.
%
I will explain the functioning of the module based on the test case "sites2".
%
This test uses a trivial polypeptide with only two titratable sites,
histidine and glutamate.
%
The test "histidine" uses only one site. The other tests use real-life,
although small proteins.


By default, the \modulename module will not empty the scratch directory
after completing the run.
%
Starting the job for the second time will cause reading of the existing 
scratch files instead of running the calculations anew.
%
To avoid this behavior, remove the scratch directory or declare another one
during the setup of the continuum electrostatic model.

\textbf{Notice! The energies calculated by the \modulename module are 
given in kcal/mol, in contrast to kJ/mol, which are normally used 
in pDynamo.}


%-------------------------------------------------
\subsection{Setup of the protein model}
The electrostatic model used by the \modulename module requires that the protein
of interest is described by the CHARMM energy model\cite{MacKerell1998}.
%
In the first step, prepare CHARMM topology (PSF) and coordinate (CRD) files.
%
The preparation can be done using the programs CHARMM\cite{CHARMM_Brooks1983}
or VMD\cite{VMD1996}.
%
During the preparation of the protein model, all titratable residues in the protein
should be set to their standard protonation states at pH = 7, i.e. aspartates
and glutamates deprotonated, histidines doubly protonated, other residues
protonated.
%
The topology, coordinate and parameter files are loaded at the
beginning of the script \texttt{sites2.py}:

{\footnotesize \begin{lstlisting}
par_tab = ["charmm/toppar/par_all27_prot_na.inp", ]
mol  = CHARMMPSFFile_ToSystem ("charmm/testpeptide_xplor.psf", isXPLOR = True, parameters = CHARMMParameterFiles_ToParameters (par_tab))
mol.coordinates3 = CHARMMCRDFile_ToCoordinates3 ("charmm/testpeptide.crd")
\end{lstlisting} }


%-------------------------------------------------
\subsection{Setup of the continuum electrostatic model}
In the second step, a continuum electrostatic model is created:

{\footnotesize \begin{lstlisting}
ce_model = MEADModel (meadPath = "/home/mikolaj/local/bin/", gmctPath = "/home/mikolaj/local/bin/", scratch = "scratch", nthreads = 2)
\end{lstlisting} }

\bigskip
The parameter "meadPath" tells the directory where the MEAD programs,
\texttt{my\_2diel\_solver} and \texttt{my\_3diel\_solver}, are located.
%
The parameter "gmctPath" is the location of the \texttt{gmct} program.
%
If none of these directories are given, \texttt{/usr/bin} is assumed
by default.
%
The parameter "scratch" tells the directory where the MEAD job files
and output files will be written to.
%
If not present, this directory will be created.
%
The last parameter, "nthreads", defines the number of threads to be used.
%
By default \texttt{nthreads=1}, which means serial run.
%
Note that "nthreads" can be any natural number and that the calculations
scale linearly with the number of threads.
%
Parallelization is done at the coarse-grain level.
%
Since the electrostatic energy terms for a particular instance of a titratable
site can be calculated independently from energy terms of other instances of other
sites, each instance is assigned a separate thread.
%
A similar approach is taken during the calculations of titration curves,
where each pH-step of a curve is calculated separately.


\bigskip
In the next step, the continuum electrostatic model is initialized:

{\footnotesize \begin{lstlisting}
ce_model.Initialize (mol)
\end{lstlisting} }

\bigskip
The initialization means partitioning of the protein into titratable sites and
a non-titratable background.
%
It also means generating model compounds.
%
At this point, however, the input files for MEAD are not written and only
the necessary data structures inside the \texttt{MEADModel} object are created.
%
The \texttt{Initialize} method takes at least one argument,
which indicates the CHARMM-based protein model.

\bigskip
The next two lines generate a summary of the continuum electrostatic model and
write a table of titratable sites:

{\footnotesize \begin{lstlisting}
ce_model.Summary ()
ce_model.SummarySites ()
\end{lstlisting} }

\bigskip
After the model has been initialized, the input files necessary for calculations in MEAD
can be written to the scratch directory:

{\footnotesize \begin{lstlisting}
ce_model.WriteJobFiles (mol)
\end{lstlisting} }

By default, each site is assigned a separate directory, for example
\texttt{scratch/PRTA/GLU8}.
%
Inside the directory, there are PQR files for each instance of the site in the protein
and in a model compound.
%
The OGM and MGM files specify parameters of lattices used for solving the Poisson-Boltzmann equation.
%
The \texttt{back.pqr} file defines the non-titratable background.
%
The \texttt{protein.pqr} file defines the whole protein and is used to calculate the boundary
between the protein and the solvent.
%
The last file, \texttt{sites.fpt}, contains atomic coordinates and charges of all instances
and is used for calculating interaction energies between
different instances of sites in the protein.


%-------------------------------------------------
\subsection{Calculating electrostatic energies}
At this point, the electrostatic energy terms can be calculated:

{\footnotesize \begin{lstlisting}
ce_model.CalculateElectrostaticEnergies ()
\end{lstlisting} }

For each instance of each site, two electrostatic energy terms are calculated, namely
the Born energy ($G_{\mathrm{Born}}$) and the background energy ($G_{\mathrm{back}}$).
%
Born energy is the electrostatic energy of a set of charges interacting with its own
reaction field.
%
Background energy is the electrostatic energy of a set of charges interacting with
charges from outside of this set.
%
The two energies are calculated for a particular instance of a site both in the model compound
and in the protein.
%
The difference
$(G_{\mathrm{Born, protein}} + G_{\mathrm{back, protein}}) - (G_{\mathrm{Born, model}} + G_{\mathrm{back, model}})$
is calculated, which is called the heterogeneous transfer energy, $G_{\mathrm{heterotrans}}$.
%
Transferring of a site means moving it from the model compound to the protein.
%
In the model compound, the site has a model energy $G_{\mathrm{model}}$, which
corresponds to the experimentally known \pka value of the deprotonation reaction.
%
The site in the protein has an intrinsic energy $G_{\mathrm{intr}} = G_{\mathrm{model}} + G_{\mathrm{heterotrans}}$.
%
The \texttt{my\_2diel\_solver} program calculates $G_{\mathrm{Born, model}}$ and $G_{\mathrm{back, model}}$.
%
The \texttt{my\_3diel\_solver} program calculates $G_{\mathrm{Born, protein}}$ and $G_{\mathrm{back, protein}}$
and, additionally, electrostatic interaction energies of an instance of a site
with other instances of other sites in the protein.
%
The \modulename module collects 
$G_{\mathrm{Born, protein}}$, $G_{\mathrm{back, protein}}$, 
$G_{\mathrm{Born, model}}$, $G_{\mathrm{back, model}}$ and interaction energies from MEAD 
and calculates $G_{\mathrm{heterotrans}}$ and $G_{\mathrm{intr}}$ for each instance of each site.

Note that the \texttt{my\_3diel\_solver} program can in principle perform calculations
in a three-dielectric environment (solvent, protein, vacuum).
%
However, the model implemented in the \modulename module only deals with two-dielectric environments, 
i.e. the solvent phase and the protein/model compound phase.


%-------------------------------------------------
\subsection{Calculating microstate energies}

After the $G_{\mathrm{intr}}$ values and interaction energies have been calculated 
for all instances of all titratable sites, 
one can calculate the energy of a particular protonation state of the protein, i.e. 
the microstate energy, $G_{\mathrm{micro}}$.
%
The polypeptide in the "sites2" example contains only two sites, 
glutamate and histidine,
so there can be $2^1 * 4^1 = 8$ possible protonation states, because glutamate has two
instances ("p" and "d") and histidine has four instances ("HSP", "HSD", "HSE", "fully deprotonated").
%
For real-life proteins, the number of protonation states is very large.
%
The protonation state of the protein is defined by a state vector.
%
Each component of the state vector represents the protonation state of a site in the protein.
%
The value of the component indicates the current instance of the site.
%
The values of 0 and 1 do not necessarily mean "deprotonated" and "protonated".
%
The next few lines of code generate all possible permutations of the state vector and calculate
their respective microstate energies:

{\footnotesize \begin{lstlisting}
statevector = StateVector (ce_model)
increment   = True
while increment:
  Gmicro = vector.CalculateMicrostateEnergy (ce_model, pH = 7.0)
  statevector.Print (ce_model, title = "Gmicro = %f" % Gmicro)
  increment = statevector.Increment ()
\end{lstlisting} }

\bigskip
In the lowest energy protonation state, the histidine is protonated at 
positions $\delta$ and $\epsilon$ and the glutamate is deprotonated.
%
Dissociation of the $\epsilon$-hydrogen from the histidine requires only 0.7\,kcal/mol.


%-------------------------------------------------
\subsection{Calculating protonation probabilities}
% For a given site, the probability of occurrence of a particular instance is calculated 
% from the Boltzmann weighted sum, which takes the form:
One option for calculating protonation probabilities is to use GMCT externally after 
the calculations by MEAD are complete.
%
This can be done by writing two files, \texttt{gintr.dat} and \texttt{W.dat}, containing 
the calculated intrinsic energies and interaction energies, respectively.

{\footnotesize \begin{lstlisting}
ce_model.WriteGintr ()
ce_model.WriteW ()
\end{lstlisting} }


\bigskip
In the \modulename module, protonation probabilities can be calculated either analytically for
small proteins or by using GMCT for larger proteins.
%
The following two lines calculate protonation probabilities analytically and report them
in a table.
%
By default, the calculations are done at pH = 7.
%
The default behavior can be changed by passing the "pH" argument. 

{\footnotesize \begin{lstlisting}
ce_model.CalculateProbabilitiesAnalytically ()
ce_model.SummaryProbabilities ()
\end{lstlisting} }


\bigskip
For comparison, the probabilities are calculated by using GMCT. 
%
During the calculations in GMCT, a \texttt{scratch/gmct/} directory will be created
for storing input and output files.

{\footnotesize \begin{lstlisting}
ce_model.CalculateProbabilitiesGMCT ()
ce_model.SummaryProbabilities ()
\end{lstlisting} }

\bigskip
The resulting tables show for each site the probability of occurrence of each instance. 
%
Instances with the highest probability of occurrence are marked with "*".
%
Optionally, the argument \texttt{reportOnlyUnusual=True} will only show sites 
in their non-standard  protonation states, for example protonated glutamates.


%-------------------------------------------------
\subsection{Calculating titration curves}
Titration curves are obtained by calculating protonation probabilities for a given pH-range, 
usually from 0 to 14.

{\footnotesize \begin{lstlisting}
ce_model.CalculateCurves (isAnalytic = True, forceSerial = True, directory = "curves_analytic")
\end{lstlisting} }

\bigskip
The argument \texttt{isAnalytic=True} tells that the protonation probabilities should be
calculated analytically.
%
Each pH-step will be calculated in a separate thread.
%
However, the analytic evaluation of titration curves seems to be very inefficient in parallel 
mode.
%
The argument \texttt{forceSerial=True} will switch on serial mode regardless of the number 
of threads declared previously.
%
For each instance of each site, a data file containing the titration curve will be generated.
%
The files will be placed in the directory specified by the argument \texttt{directory}.
%
The curves can be plotted in a plotting program, for example Gnuplot.

\bigskip
The same task can be accomplished by using GMCT in parallel mode:

{\footnotesize \begin{lstlisting}
ce_model.CalculateCurves (directory = "curves_gmct")
\end{lstlisting} }


%-------------------------------------------------
\subsection{Calculating microstate energies of substates}
From the calculated protonation probabilities, one can generate a state vector describing
the lowest energy protonation state:

{\footnotesize \begin{lstlisting}
vector_lowest = StateVector_FromProbabilities (ce_model)
\end{lstlisting} }

\bigskip
The microstate energy ($G_{\mathrm{micro}}$) in the lowest energy protonation state can 
be calculated in the following way:

{\footnotesize \begin{lstlisting}
Gmicro = vector_lowest.CalculateMicrostateEnergy (ce_model)
\end{lstlisting} }

\bigskip
By changing the values (=instances) of selected components (=sites) of the state vector, 
one can calculate energies for a series of substates of the lowest energy state.
%
On the other hand, substate energies can be calculated easier with the \texttt{MEADSubstate} class:

{\footnotesize \begin{lstlisting}
sites = (("PRTA", "ASP",  2), ("PRTA", "GLU", 14), ("PRTA", "ARG", 15))
substate = MEADSubstate (ce_model, sites, pH = 7.0)
substate.CalculateSubstateEnergies ()
substate.Summary ()
\end{lstlisting} }

\bigskip
The above code was taken from the test case "defensin".
%
Note that \texttt{pH = 7.0} is an optional argument and that the given pH should be the same
as the pH used during the calculation of protonation probabilities.
%
The code will generate a table of substates with their corresponding relative energies:

\linespread{0.8}
{\footnotesize \begin{lstlisting}
-------------------------------------------------------------------------
 State  Gmicro  Charge  Protons  PRTA ASP 2    PRTA GLU 14   PRTA ARG 15
-------------------------------------------------------------------------
     1     0.00      -1       1      d             d             p       
     2     4.55       0       2      p             d             p       
     3     6.69      -2       0      d             d             d       
     4    10.05       0       2      d             p             p       
     5    11.21      -1       1      p             d             d       
     6    14.86       1       3      p             p             p       
     7    16.62      -1       1      d             p             d       
     8    21.40       0       2      p             p             d       
-------------------------------------------------------------------------
\end{lstlisting} }
\linespread{1.6}

\bigskip
In the lowest energy state, Asp2 and Glu14 are deprotonated, while Arg15 and Cys31
are protonated.
%
Protonating Asp2 requires about 4.6\,kcal/mol.


%-------------------------------------------------
\section{Parameter files}
Two parameter formats are recognized, 
YAML (pDynamo's default format) and EST (QMPB's format).
%
The file \texttt{radii.yaml} defines a list of atomic radii 
required to calculate the boundary between the protein phase and 
the solvent phase.
%
The files in the \texttt{sites/} directory define parameters for titratable 
sites.
%
The listing below shows the parameter file for aspartate:

\newpage

\linespread{0.8}
{\footnotesize \begin{lstlisting}
---
site      : ASP
atoms     : [CB, HB1, HB2, CG, OD1, OD2]
instances :
  - label   : p
    Gmodel  : -5.487135
    protons : 1
    charges : [-0.21,  0.09,  0.09,  0.75, -0.36, -0.36]

  - label   : d
    Gmodel  : 0.000000
    protons : 0
    charges : [-0.28,  0.09,  0.09,  0.62, -0.76, -0.76]
...
\end{lstlisting} }
\linespread{1.6}


\bigskip
The site's name is defined by the parameter \texttt{site}. 
%
\texttt{atoms} defines a list of names of atoms that constitute the site.
%
The consecutive lines define instances of the site.
%
\texttt{label} is the name of the instance, for example "p" for "protonated".
%
\texttt{Gmodel} is the energy of the model compound expressed in kcal/mol.
%
$G_{\mathrm{model}}$ has been calculated at the temperature of 300\,K from the 
\pka value of aspartate, which is 4.0.
%
The values of model energies are recalculated if a different temperature is 
to be used.
%
The reason for $G_{\mathrm{model}} = 0$ for the deprotonated instance is that 
the energies are calculated relative to a reference state, which is a fully 
deprotonated state.
%
The parameter \texttt{protons} tells the number of protons bound to the site and 
is used during the calculation of the microstate energy.
%
Finally, \texttt{charges} is a list of atomic charges that differ between 
the instances.
%
The order of charges is the same as the order of atoms in the \texttt{atoms} list.
%
The charges usually come from the CHARMM force field, i.e. they are
parametrized for the dielectric constant $\epsilon = 4$.

Proteins often contain non-standard titratable sites, for example ligands or cofactors.
%
The parameter files describing these sites are recognized by their extension and can be
loaded automatically from the current directory.
%
These additional parameter files can be provided both in EST or YAML formats (the EST format
takes precedence).
%
If the names of the new sites coincide with the names of already existing sites, the
parameters of the existing sites will be overwritten.


%=================================================
\section{Future developments}
%=================================================
\linespread{0.8}

\begin{itemize}
  \setlength{\itemsep}{1.0pt}
  \item Extend the model by including rotameric instances
  \item Include quantum chemical sites as in QMPB
  \item Include reduction/oxidation reactions in addition to protonation/deprotonation reactions
  \item Enclose the MEAD and GMCT codes inside pDynamo
  \item Optionally convert kcal/mol (MEAD units) to kJ/mol (pDynamo units)
  \item Improve the efficiency of writing job files
  \item Make use of the \texttt{pqr2SolvAccVol} program to speed up calculations 
\end{itemize}

\linespread{1.6}


%=================================================
\section{Feedback}
%=================================================
The \modulename module is in a stage of active development.
%
Therefore, bugs are inevitable.

If you found a bug, have a question or wish to comment or contribute to the code, 
please write to 
\makebox{\href{mailto:mikolaj.feliks@gmail.com}{mikolaj.feliks@gmail.com}}.
%
Alternatively, you can post comments or report problems on the project's website,
\url{http://code.google.com/p/pdynamo-extensions/}.


%=================================================
\section{References}
%=================================================
\renewcommand{\refname}{}
\vspace*{-1cm}

\linespread{0.8}
\bibliographystyle{ieeetr}
\bibliography{refs}

\linespread{1.6}


%-------------------------------------------------
\subsection*{Websites}

\begin{itemize}
  \item pDynamo (Martin J. Field)\\ \url{http://pdynamo.org}

  \item MEAD (Donald Bashford)\\ \url{http://stjuderesearch.org/site/lab/bashford/}

  \item Extended MEAD (Thomas Ullmann)\\ \url{http://www.bisb.uni-bayreuth.de/People/ullmannt/index.php?name=extended-mead}

  \item GMCT (Thomas Ullmann)\\ \url{http://www.bisb.uni-bayreuth.de/People/ullmannt/index.php?name=gmct-gcem}

  \item Doctoral thesis of Timm Essigke (which was my primary source of knowledge)\\ \url{https://epub.uni-bayreuth.de/655/}

% Talk about most informative chapters?

  \item Doctoral thesis of Thomas Ullmann\\ \url{https://epub.uni-bayreuth.de/199/}

  \item Doctoral thesis of Astrid Klingen\\ \url{https://epub.uni-bayreuth.de/763/}
\end{itemize}


\end{document}
